Generalized Bradley-Terry models

### 2025-03-13 inst/jss_hyper3_real.Rnw
label final node
slight rewriting

  A generalization of \code{hyper2} is presented in which
  Bradley-Terry strengths may be modified using a multiplicative
  factor $\lambda$; Generally, ordinary Bradley-Terry is recovered if
  $\lambda=1$.  This approach offers multiple advantages over other
  generalizations such as reified Bradley-Terry.  It becomes possible
  to represent non-independence of successive observations in
  Plackett-Luce likelihood functions which allows a parametrization of
  the ``red bus, blue bus'' problem in preference analysis.  I present
  examples drawn from the fields of food preference and the e-sport
  {\em Counterstrike}.  }

@@

The Bradley-Terry model for datasets involving paired comparisons has
wide uptake in the {R} community.  The canonical case is to
consider~$n$ players who compete against one another in pairs; the
basic inference problem is to estimate
numbers~$\mathbf{p}=\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$,
$\sum p_i=1$ so that
$\Prob(\mbox{$i$~beats~$j$})=\frac{p_i}{p_i+p_j}$.  Information about
the~$p_i$ may be obtained from the results of paired comparisons
between the players.

@@

where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special move, and $\mu$ the enhancement afforded to $p_4$ by
the drug; values $>1$ would correspond to a positive effect.  In this
paper I generalize the {\tt hyper2} suite for Bradley-Terry
likelihoods~\citep{hankin2017_rnw} to accommodate this class of
multiplicative factors.  The most general likelihood function would be
proportional to

@@

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{1mm}{6mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{1mm}{6mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 669 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt soccerworld}}
\end{table}

@@

where the home-ground advantage is represented by~$\lambda$.  In {\tt
  hyper3} idiom, a term such as $\lambda p_R + p_C$ is represented as
a named numeric vector.  For example, if the home strength were
$\lambda = 1.8$ then the first bracket would be {\tt c(Rangers=1.8,
  Celtic=1)} with an associated power of $-81$.  The computational
methodology is a generalization of that used
in~\cite{hankin2024_hyper3} in which the multiplicative factors
correspond to number of clones.  However, in~\cite{hankin2024_hyper3}
the factors are known non-negative integers, and here they are, in
general, unknown non-negative real numbers.

@@

\fill (10.6, 6) circle[radius=2pt];  % abc finishes; paths de
\node at (10.6,5.8) {$c$};
\draw[thick] (10.6, 6) -- (11.2,7) -- (12.2,7);
\fill (12.2,7) circle[radius=2pt];
\node[anchor = west] at (12.2,7) {$a\succ b\succ c\succ d\succ e$};
\node at (11.8, 7.4) {$\frac{p_d}{p_d+p_e}$};

\draw[thick] (10.6, 6) -- (11.2,6.2) -- (12.2,6.2);
\fill (12.2,6.2) circle[radius=2pt];
\node[anchor = west] at (12.2,6.2) {$a\succ b\succ c\succ e\succ d$};
\node at (11.8, 6.6) {$\frac{p_e}{p_d+p_e}$};

@@

\draw[thick] (10.6, 5) -- (11.2,5.3) -- (12.2, 5.3);
\fill (12.2,5.3) circle[radius=2pt];
\node[anchor = west] at (12.2,5.3) {$a\succ b\succ d\succ c\succ e$};
\node at (11.8, 5.7) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};

\draw[thick] (10.6, 5) -- (11.2,4.7) -- (12.2, 4.7);
\fill (12.2,4.7) circle[radius=2pt];
\node[anchor = west] at (12.2,4.7) {$a\succ b\succ d\succ e\succ c$};
\node at (11.8, 5.0) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};

\fill (10.6, 4) circle[radius=2pt];
\node at (10.6,3.8) {$e$};
\draw[thick] (10.6, 4) -- (11.2,3.9) -- (12.2, 3.9);
\fill (12.2,3.9) circle[radius=2pt];
\node[anchor = west] at (12.2,3.9) {$a\succ b\succ e\succ c\succ d$};
\node at (11.8, 4.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_d}}$};

\draw[thick] (10.6, 4) -- (11.2,3.1) -- (12.2,3.1);
\fill (12.2,3.1) circle[radius=2pt];
\node at (11.8, 3.5) {$\frac{\ld{p_d}}{\la{p_c}+\ld{p_d}}$};
\node[anchor = west] at (12.2,3.1) {$a\succ b\succ e\succ d\succ c$};

\draw[thick] (4.8, 0) -- (5.8,1) -- (8.2,1);
\fill (12.2,1) circle[radius=2pt];
\node[anchor = west] at (12.2,1) {$d\succ a\succ c\succ e\succ b$};
\node at (7, 1.4)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};

\draw[thick] (4.8, 0) -- (5.8,0) -- (8.2,0);
\fill (12.2,0) circle[radius=2pt];
\node[anchor = west] at (12.2,0) {$d\succ e\succ a\succ b\succ c$};
\node at (7, 0.4)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};

@@

\fill (12.2,-1) circle[radius=2pt];
\node[anchor = west] at (12.2,-1) {$d\succ e\succ a\succ c\succ b$};

@@

\draw[thick](10.6, -2) -- (11.2,-1.8) -- (12.2,-1.8);
\fill (12.2,-1.8) circle[radius=2pt];
\node[anchor = west] at (12.2,-1.8) {$d\succ e\succ b\succ a\succ c$};
\node at (11.8, -1.4) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};

\draw[thick] (10.6, -2) -- (11.2,-2.6) -- (12.2,-2.6);
\fill (12.2,-2.6) circle[radius=2pt];
\node[anchor = west] at (12.2,-2.6) {$d\succ e\succ b\succ c\succ a$};
\node at (11.8, -2.2) {$\frac{\la{p_c}}{\la{p_a}+\la{p_c}}$};

@@

\draw[thick] (10.6, -3) -- (11.2,-3.4) -- (12.2, -3.4);
\fill (12.2,-3.4) circle[radius=2pt];
\node[anchor = west] at (12.2,-3.4) {$d\succ e\succ c\succ b\succ a$};

@@

\draw[thick] (10.6, 1) -- (11.2,2) -- (12.2,2);
\node at (11.8, 2.4) {$\frac{\la{p_b}}{\la{p_b}+\ld{p_e}}$};
\fill (12.2,2) circle[radius=2pt];
\node[anchor = west] at (12.2,2) {$d\succ a\succ c\succ b\succ e$};
\draw[thick] (10.6, 1) -- (11.2,1) -- (12.2,1);
\node at (11.8, 1.4) {$\frac{\ld{p_e}}{\la{p_b}+p_e}$};

@@

Then, as before, we may define a profile likelihood function of
$\lambda$, Figure~\ref{profilesupportsushi}.

@@

<<findstuffsushi,echo=FALSE,cache=TRUE>>=
osushi <- optimize(f, c(2,4), maximum=TRUE)
@

<<tinysushi,echo=FALSE>>=
sushi_maxlambda <- osushi$maximum
@

@@

\begin{figure}[htbp]
  \begin{center}
<<plotlikesushi, echo=FALSE>>=
plot(log(s),L-max(L),type='b',
     xlab=expression(log(lambda)), ylab='Log-likelihood')
abline(v=0,col='gray')
abline(h=c(0,-2),col='gray')
segments(x0 = log(sushi_maxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(sushi_upper    ), y0 = -2.2, y1 = -1.8, col = 'blue')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$ in the case of sushi preferences.
  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
  show the two-units-of-support limits with the credible interval of
  $(\Sexpr{round(sushi_lower,2)},\Sexpr{round(sushi_upper,2)})$,
  marked in blue.  Red bar shows evaluate at
  $\lambda\simeq\Sexpr{round(sushi_maxlambda,2)}$}
\end{center}
\end{figure}

>>>>

  A generalization of the Bradley-Terry model is presented in which
  strengths may be modified using a multiplicative factor $\lambda$;
  Generally, ordinary Bradley-Terry is recovered if $\lambda=1$.  This
  approach offers multiple advantages over other generalizations such
  as reified Bradley-Terry.  It becomes possible to represent
  non-independence of successive observations in Plackett-Luce
  likelihood functions which allows a parametrization of the ``red
  bus, blue bus'' problem in preference analysis.  I present examples
  drawn from the fields of food preference and the e-sport {\em
    Counterstrike}.  }

@@

The Bradley-Terry model for datasets involving paired comparisons has
wide uptake in the {R} community.  The canonical case is to
consider~$n$ players who compete against one another in pairs; the
basic inference problem is to estimate
numbers~$\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$, $\sum
p_i=1$ so that $\Prob(\mbox{$i$~beats~$j$})=\frac{p_i}{p_i+p_j}$.
Information about the~$p_i$ may be obtained from the results of paired
comparisons between the players.

@@

where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special move, and $\mu$ the enhancement afforded to $p_4$ by
the drug; values $>1$ would correspond to a positive effect.  In this
paper I generalize the {\tt hyper2} R package~\citep{hankin2017_rnw},
available on CRAN~\citep{rcore2025}, for Bradley-Terry likelihoods to
accommodate this class of multiplicative factors.  The most general
likelihood function would be proportional to

@@

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 669 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt soccerworld}}
\end{table}

@@

where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.

@@

\fill (10.6, 6) circle[radius=2pt];  % abc finishes; paths de
\node at (10.6,5.8) {$c$};
\draw[thick] (10.6, 6) -- (11.2,7) -- (12.2,7);
\fill (12.2,7) circle[radius=2pt];
\node at (12.3,6.8) {$d$};
\node[anchor = west] at (12.5,7) {$a\succ b\succ c\succ d\succ e$};
\node at (11.8, 7.4) {$\frac{p_d}{p_d+p_e}$};

\draw[thick] (10.6, 6) -- (11.2,6.2) -- (12.2,6.2);
\fill (12.2,6.2) circle[radius=2pt];
\node at (12.3,6.0) {$e$};
\node[anchor = west] at (12.5,6.2) {$a\succ b\succ c\succ e\succ d$};
\node at (11.8, 6.6) {$\frac{p_e}{p_d+p_e}$};

@@

\draw[thick] (10.6, 5) -- (11.2,5.3) -- (12.2, 5.3);
\fill (12.2,5.3) circle[radius=2pt];
\node at (12.3,5.1) {$c$};
\node[anchor = west] at (12.5,5.3) {$a\succ b\succ d\succ c\succ e$};
\node at (11.8, 5.7) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};

\draw[thick] (10.6, 5) -- (11.2,4.7) -- (12.2, 4.7);
\fill (12.2,4.7) circle[radius=2pt];
\node at (12.3,4.5) {$e$};
\node[anchor = west] at (12.5,4.7) {$a\succ b\succ d\succ e\succ c$};
\node at (11.8, 5.0) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};

\fill (10.6, 4) circle[radius=2pt];
\node at (10.6,3.8) {$e$};
\draw[thick] (10.6, 4) -- (11.2,3.9) -- (12.2, 3.9);
\fill (12.2,3.9) circle[radius=2pt];
\node at (12.3,3.7) {$c$};
\node[anchor = west] at (12.5,3.9) {$a\succ b\succ e\succ c\succ d$};
\node at (11.8, 4.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_d}}$};

\draw[thick] (10.6, 4) -- (11.2,3.1) -- (12.2,3.1);
\fill (12.2,3.1) circle[radius=2pt];
\node at (11.8, 3.5) {$\frac{\ld{p_d}}{\la{p_c}+\ld{p_d}}$};
\node at (12.3,2.9) {$d$};
\node[anchor = west] at (12.5,3.1) {$a\succ b\succ e\succ d\succ c$};

\draw[thick] (4.8, 0) -- (5.8,1) -- (8.2,1);
\fill (12.2,1) circle[radius=2pt];
\node at (12.3,0.8) {$e$};
\node[anchor = west] at (12.5,1.0) {$d\succ a\succ c\succ e\succ b$};
\node at (7, 1.4)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};

\draw[thick] (4.8, 0) -- (5.8,0) -- (8.2,0);
\fill (12.2,0) circle[radius=2pt];
\node at (12.3, -0.2) {$b$};
\node[anchor = west] at (12.5, 0) {$d\succ e\succ a\succ b\succ c$};
\node at (7, 0.4)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};

@@

\fill (12.2,-1) circle[radius=2pt];
\node at (12.3,-1.2) {$c$};
\node[anchor = west] at (12.5,-1) {$d\succ e\succ a\succ c\succ b$};

@@

\draw[thick](10.6, -2) -- (11.2,-1.8) -- (12.2,-1.8);
\fill (12.2,-1.8) circle[radius=2pt];
\node at (12.3, -2.0) {$a$};
\node[anchor = west] at (12.5, -1.8) {$d\succ e\succ b\succ a\succ c$};
\node at (11.8, -1.4) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};

\draw[thick] (10.6, -2) -- (11.2,-2.6) -- (12.2,-2.6);
\fill (12.2,-2.6) circle[radius=2pt];
\node at (12.3, -2.8) {$c$};
\node[anchor = west] at (12.5,-2.6) {$d\succ e\succ b\succ c\succ a$};
\node at (11.8, -2.2) {$\frac{\la{p_c}}{\la{p_a}+\la{p_c}}$};

@@

\draw[thick] (10.6, -3) -- (11.2,-3.4) -- (12.2, -3.4);
\fill (12.2,-3.4) circle[radius=2pt];
\node at (12.3, -3.6) {$b$};
\node[anchor = west] at (12.5,-3.4) {$d\succ e\succ c\succ b\succ a$};

@@

\draw[thick] (10.6, 1) -- (11.2,2) -- (12.2,2);
\node at (11.8, 2.4) {$\frac{\la{p_b}}{\la{p_b}+\ld{p_e}}$};
\fill (12.2,2) circle[radius=2pt];
\node at (12.3, 1.8) {$b$};
\node[anchor = west] at (12.5, 2) {$d\succ a\succ c\succ b\succ e$};
\draw[thick] (10.6, 1) -- (11.2,1) -- (12.2,1);
\node at (11.8, 1.4) {$\frac{\ld{p_e}}{\la{p_b}+p_e}$};

@@

@@

% chunk 'findstuffsushi' takes about 10-20 minutes to run

<<findstuffsushi,echo=FALSE, cache=TRUE>>=
osushi <- optimize(f, c(2,4), maximum=TRUE)
@

% chunk tinysushi' takes about 5 minutes to run

<<tinysushi, echo=FALSE, cache=TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@

% chunk 'findotherstuffsushi' takes about 10 minutes to run

@@

(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$-value of about
$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.

\begin{figure}[htbp]
  \begin{center}
<<plotlikesushi, echo=FALSE>>=
plot(log(s),L-max(L),type='b',
     xlab=expression(log(lambda)), ylab='Log-likelihood')
abline(v=0,col='gray')
abline(h=c(0,-2),col='gray')
segments(x0 = log(sushimaxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(sushi_upper    ), y0 = -2.2, y1 = -1.8, col = 'blue')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$ in the case of sushi preferences.
  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
  show the two-units-of-support limits with the credible interval of
  $(\Sexpr{round(sushi_lower,2)},\Sexpr{round(sushi_upper,2)})$,
  marked in blue.  Red bar shows evaluate at
  $\lambda\simeq\Sexpr{round(sushimaxlambda,2)}$}
\end{center}
\end{figure}
